Military mechanization as a complex system

### 2018-07-14 paper2/paper2.tex
werk

\usepackage{amsmath}
\usepackage{graphicx}
\graphicspath{}

@@

\title{An agent-based model of military mechanization}

@@

	Military mechanization is the degree to which a state's army is comprised of
	ground combat vehicles as opposed to infantry. Mechanization is both a 
	theoretically important topic in international relations scholarship and a practical 
	concern for policymakers. \citet{sechser2010army} influentially argue that
	inter-state dynamics (e.g., conflict) better explain mechanization choices than do
	intra-state factors (e.g., culture). This paper demonstrates that
	the multivariate statistical approach employed in \citet{sechser2010army} is
	inadequate and that agent-based modeling is needed to capture
	international security competition as a complex system. Candidate
	agent-based models are presented.

@@

@@

Maybe it's because states are not unitary. Not necessariy requirement: e.g.,
household is agent in Schelling model.

Cite Cederman to say why I don't do inheritenace - reification of actor

Why ABM:
 - inadequacy of SS - VIF, linear model
 - uncertainty of causation
 - endogeneity
 - GDP / capita explains this...but also everything else
 - have to go with simulation
 - bigger picture, world politics as complex system
 - look at Axtell \citep{axtell2000agents}

Things they don't capture: selection into war and deterrence, gdp / capita to
political structure, institutional culture and choice, network effects and
emulation

Two central claims of this paper are that the multivariate statistical approach
of \citet{sechser2010army} is inadequate and agent-based modeling is needed.
They are addressed in turn.

\citet{sechser2010army} gather observational panel data that consists of a measure 
of military mechanization $y$ indexed by state $i$ and observation time $k$, as well as
time-indexed covariates $x$. The association of covariates with mechanization is
estimated via ordinary least squares (OLS) linear regression:

\begin{equation}
	y_{ik} = \beta_0 + \beta_j x_{ijk} + \epsilon \hspace{1em}
\end{equation}

Multicollinearity - everything correlated with each other.

The fitted model attempts to characterize the separate
effect of each factor upon mechanization. However, evidence suggests that
the covariates in fact are strongly cross-correlated. The problem of
multi-collinearity can be expressed both

This paper presents an agent-based model of military mechanization. It
reviews the academic literature on military mechanization. It discusses why
orthodox modeling strategies fail to capture international security competition
as a complex system and why agent-based modeling is needed. In support of these
claims, it ``agentizes'' the leading model of military mechanization and shows that the
results of doing cast doubt on it. Finally, it presents a
novel agent-based model that better represents military mechanization.

In summary, the orthodox view of mechanization is that it is a
reflexive process: states choose their mechanization posture largely based on
that neighbors, enemies, and recent military experiences (e.g. defeat in
counter-insurgency). This paper argues
that this view---based on traditional longitudinal data analysis---is ripe for
agentization. It implicitly argues for autonomous agents (states) that have
cognition regarding mechanization that takes into account other agents (e.g.
enemies), spatiality (e.g. neighbors), and learning from experience. This paper
builds such a model and validates it against the empirical record, using the
National Mechanization Index of \citet{sechser2010army}. It finds that the 
orthodox view of mechanization does not stand up to agentization. A positive feedback loop 
causes states to
mechanize much faster and to a greater degree than occurs in the real-world.
This implies the existence of some countervailing factor (possibly a negative
feedback loop) that mitigates the extent of mechanization. Candidates for what
this could be mostly center on ``internal'' factors---e.g., culture, economy,
and so on---as opposed to the ``external'', security environment-centric
orthodox view.

There are many candidates for the causes of mechanization. Inter-state
competition and war is one. States may choose choose mechanization policies 
that they believe will allow them to prevail against potential adversaries, or to send a
signal of deterrence, or to learn from the perceived mistakes of the last war
\citep{mearsheimer1983conventional,huth1988extended,murray2011military}. 
Economic endowments may play a role: e.g., capital-rich states may mechanize more
than labor-rich states \citep{gartzke2001democracy}. Domestic political
institutions may matter: e.g., autocracies may be more willing than democracies to adopt
low-mechanization ``human wave'' tactics
\citep{quinlivan1999coup,reiter2002democracies,brooks2008shaping,talmadge2015dictator}. 
Ideology \citep{van1984cult} and national culture \citep[e.g.][]{pollack2004arabs} are
hard-to-measure but perhaps influential. Lastly,
organizational culture can play a role \citep{builder1989masks}. Militaries may pursue a
certain mechanization posture because it aligns with their institutional
objectives, which may or may not be coincident with rational strategy. 
Consider Ireland: between the 1920s and 1940s, it seriously considered expending a 
large portion of its defense budget on building a relatively small number of tanks. 
They would have been wholly insufficient to repel a British invasion (their ostensible purpose), 
and Ireland would have been better-served to pursue a low-mechanization guerrilla defense strategy, 
but tanks were seen as a prestigious signifier of a ``real'' professional
military and hence desirable
\citep{farrell1998professionalization,farrell2001transnational}.

\citet{sechser2010army} is the landmark study on military mechanization. They
assemble a longitudinal dataset at the country-year level of the mechanization
of all states between 1979 and 2001. They conduct regression analysis to identify
the effect of covariates on mechanization. They conclude that out of the
aforementioned potential causes---security environment, economy, domestic
politics, and ideology / culture---that the first is most important: ``choices about
mechanization are strongly associated with a state's security environment.''

\subsection{The case for agent-based modeling}

There is reason to believe that the orthodox multivariate statistical modeling 
approach employed in \citet{sechser2010army} is insufficient. Several factors point towards 
agent-based modeling being a more appropriate choice.

One, from a stat

>>>>

\usepackage{amsmath}
\usepackage{graphicx}
\graphicspath{}
\usepackage{tikz}

@@

\title{Military mechanization as a complex system}

@@

	Military mechanization is the degree to which a state's army is comprised of
	ground combat vehicles as opposed to infantry. Mechanization is both a 
	theoretically important topic in international relations scholarship and a practical 
	concern for policymakers. \citet{sechser2010army} influentially argue that
	inter-state dynamics (e.g., conflict) better explain mechanization choices than do
	intra-state factors (e.g., culture).

@@

This paper presents an agent-based model of military mechanization. It
reviews the academic literature on military mechanization. It discusses why
orthodox modeling strategies fail to fully describe international security competition
as a complex system and why agent-based modeling is needed. In support of these
claims, it ``agentizes'' the leading model of military mechanization and shows
that it does not hold. Finally, it presents a
novel agent-based model that better represents military mechanization.

There are many potential causes of mechanization. Inter-state
competition and war is a prominent candidate. States may choose choose mechanization policies 
that they believe will allow them to prevail against potential adversaries, or to send a
signal of deterrence, or to learn from the perceived mistakes of the last war
\citep{mearsheimer1983conventional,huth1988extended,murray2011military}. 
Economic endowments may play a role: e.g., capital-rich states may mechanize more
than labor-rich states \citep{gartzke2001democracy}. Domestic political
institutions may matter: e.g., autocracies may be more willing than democracies to adopt
low-mechanization ``human wave'' tactics
\citep{quinlivan1999coup,reiter2002democracies,brooks2008shaping,talmadge2015dictator}. 
Ideology \citep{van1984cult} and national culture \citep[e.g.][]{pollack2004arabs} are
hard-to-measure but perhaps influential. Lastly,
organizational culture can play a role \citep{builder1989masks}. Militaries may pursue a
certain mechanization posture because it aligns with their institutional
objectives, which may or may not be coincident with rational strategy. 
Consider Ireland: between the 1920s and 1940s, it seriously considered expending a 
large portion of its defense budget on building a relatively small number of tanks. 
They would have been wholly insufficient to repel a British invasion (their ostensible purpose), 
and Ireland would have been better-served to pursue a low-mechanization guerrilla defense strategy, 
but tanks were seen as a prestigious signifier of a ``real'' professional
military and hence desirable
\citep{farrell1998professionalization,farrell2001transnational}.

\citet{sechser2010army} is the landmark study on military mechanization. They
assemble a longitudinal dataset at the country-year level of the mechanization
of all states between 1979 and 2001. They conduct regression analysis on this
panel data. They use a ordinary least squares (OLS) linear regression model with corrected standard
errors,

\begin{equation}
y_{ij} = \beta_0 + \sum_{k=1}^P \beta_{ijk}x_{ijk} + \epsilon
\end{equation}

with $y_{ij}$ the mechanization level of state $i$ in year $j$ (measured as the 
ratio of ground combat vehicles to infantry in a state's army) and $x_1 \ldots
x_j \ldots x_{P}$ covariates hypothesized to be associated with
mechanization. They conclude that, out of the
aforementioned potential causes, the first mentioned is most important: ``choices about
mechanization are strongly associated with a state's security environment.''

\section{System complexity and agent-based modeling}

Although \citet{sechser2010army} is a valuable contribution to the literature,
its orthodox multivariate statistical modeling is insufficient. Mechanization is a complex system. 
Agent-based modeling is a more appropriate paradigm for studying it. These three claims are addressed in turn.

\subsection{Inadequacy of statistical modeling}

\citet{sechser2010army} operate on observational rather than experimental
data. The assignment of treatments to subjects is not randomized, causing
\textit{selection bias}. For example, \citet{sechser2010army} operationalize
dyadic ``rivalry'' as whether two states recently engaged in a Militarized
Interstate Dispute (MID) \citep{cow_midb}, and find significant association with mechanization. 
But interstate conflict is not a randomly assigned treatment. Rather, at least one state must select
themselves into it \citep{fearon2002selection}. States differ in their
proclivity for conflict initiation based on many distinct factors. 
Thus, correlation attributed to rivalry is in fact
bundling together many uncontrolled-for covariates, casting doubt on the claimed
finding. Natural experiments---a situation in which assignation of treatment is
known to be random---are the typical econometric strategy to avoid bias. For
example, \citet{angrist2002vouchers} exploit the fact that private school
vouchers were distributed by lottery in Colombia to obtain unbiased estimates of
the effect of schooling type (private vs. public) on educational achievement.
But such situations are rare in international relations---e.g., random assignation of war is an
absurd proposition---and so
\citet{sechser2010army} lack such a control. Given rife selection effects and few
natural experiments, econometric modeling alone is an insufficient tool for studying mechanization.

\subsection{System complexity}

\subsubsection{Decomposibility}

The additive linear model in (1) assumes that each covariate $x$ is separable.  
Interpretation of the corresponding coefficient $\beta$ requires an assumption of independence 
(i.e. a lack of multicollinearity). It also makes an implicit claim about the direction of
causation: that mechanization $y$ is a function of those 
$x$.\footnote{To be clear,
\citet{sechser2010army} studiously avoid claiming causation in their text. But
their choice in what is modeled as $y$ and what is modeled $x$ says a lot about the perceived
direction of association.} These assumptions do not hold.

Let us take as an example GDP per capita. \citet{sechser2010army} find it negatively correlated with
mechanization at the $p < 0.05$ level. Unbiased estimation of this individual coefficient
requires lack of correlation between GDP per capita and other covariates, either individually or
jointly. Put differently, \textit{it requires that the system can be neatly decomposed into constituent parts}. Figure 1 is a
causal diagram depicting this logic. 
\begin{figure}[h]
	\centering
	\caption{Decomposable system}
	\begin{tikzpicture}
		\node[draw, rectangle](Mechanization)  at  (0,0) {Mechanization};	
		\node[draw, rectangle](GDP per capita)  at  (4,0) {GDP per capita};
		\node[draw, rectangle](Other covariates)  at  (4,-2) {Other covariates};
		\draw[->] (GDP per capita) -- (Mechanization);
		\draw[->] (Other covariates) -- (Mechanization);
	\end{tikzpicture}
\end{figure}

By standard statistical tests, the assumption of decomposability appears supported. The model's variance
inflation factor generally does not rise above 2, indicating a lack of
multicollinearity.\footnote{The high correlation between dummy variables}

However, the literature suggests that (1) is not a correctly specified model. GDP per capita is associated with
regime type (e.g., democracy vs. autocracy), and causality appears to flow in both directions, i.e. democracy
causes wealth and wealth also causes democracy \citep{helliwell1994empirical,barro1996democracy,robinson2006economic,doucouliagos2008democracy}. 
Regime type affects the probability of war: the closest thing to an empirical law in international relations is
that democracies do not fight each other
\citep{levy1988domestic}, while some kinds of autocracies are war-prone \citep{de1995war}. Regime type is widely argued to influence
battlefield performance in war because it affects the quality of military institutions and hence
capability of fielded forces \citep{reiter2002democracies,weeks2014dictators,talmadge2015dictator}. 
Battlefield performance affects regime type, e.g. a losing side may be subject to foreign-imposed
regime change a la 1939 France or 2003 Iraq
\citep{werner1996absolute,lo2008ensuring,downes2013forced}. \citet{sechser2010army} claim that the observed
historical probability of conflict (``rivalry'', in their terminology) affects mechanization; and claim that observed
battlefield performance (``lessons'') also affect mechanization. Mechanization surely affects the
probability of war, as it sends some kind of deterrent signal. Mechanization also surely affects
battlefield performance in war (if it did not, we would not care about it).
Mechanization also may affect GDP per capita in that states may over-spend on defense and 
impoverish the country, with the Soviet Union as a prominent example. Deterrence theory requires that 
battlefield performance affect the probability of war, i.e. militarily 
potent states are unlikely to be attacked. The exact truth value of these claimed relationships is
uncertain---in no small part due to the limitations of econometrics that were just addressed---but
they are generally uncontroversial statements in their respective fields.

\begin{figure}[h]
	\centering
	\caption{Complex system}
	\begin{tikzpicture}
		\node[draw, rectangle](Mechanization)            at  (0,0)  {Mechanization};	
		\node[draw, rectangle](GDP per capita)           at  (4,0)  {GDP per capita};
		\node[draw, rectangle](Democracy)                at  (8,0) {Democracy};
		\node[draw, rectangle](Probability of war)       at  (0,-4)  {Probability of war};	
		\node[draw, rectangle](Battlefield performance)  at  (8,-4)  {Battlefield performance};	
		\draw[->] (Mechanization) to (GDP per capita);
		\draw[->] (GDP per capita) to (Democracy);
		\draw[->] (Democracy) to (Battlefield performance);
		\draw[->] (Battlefield performance) to (Probability of war);
		\draw[->] (Probability of war) to (Mechanization);
		\draw[->] (GDP per capita) to[out=90,in=90] (Mechanization);
		\draw[->] (Democracy) to[out=270,in=270] (GDP per capita);
		\draw[->] (Battlefield performance) to[out=360,in=360] (Democracy);
		\draw[->] (Mechanization) to (Battlefield performance);
		\draw[->] (Mechanization) to[out=180,in=180] (Probability of war);
		\draw[->] (Democracy) to (Probability of war);
	\end{tikzpicture}
\end{figure}

Figure 2 is the logic diagram resulting from the above. It must be emphasized that we are only scratching the 
surface of system complexity. For example, Figure 2 does not address the nature of relationships 
between covariates: linear or non-linear? subject to multi-way
interactions? But even as-is, Figure 2 visually conveys the essential point: \textit{the system is
not tractable to decomposition}. Ascertaining cause and effect in this system through simple observation is 
near-impossible. 
One instead must perform ``causal surgery'' \citep{pearl2009causality}, i.e. modify the value at one
of the nodes and observe the
resultant system state. But, as previously covered, doing so in international relations is an
absurd proposition: one cannot mete out international conflict in a real-world experimental
setting to assess the effect on mechanization. A modeling paradigm that would let us do so is needed.

\subsubsection{Cognition}

There is no law of nature dictating that an increase in one state's mechanization causes an adversary to 
raise its own mechanization. It also cannot be claimed that many individual choices aggregate into something 
akin to natural law---the sort of logic used 
in economics to justify treating supply and demand in a physics-esque manner---because the number of
people that decide mechanization choices is small. In a personalistic
dictatorship, it might be $n=1$; even in polities with more dispersed authority, such decisions
typically occur in the rarified and thinly populated realm of generals and cabinet ministers. 
\textit{A state's mechanization level changes because someone decides to change it}.

If one takes the previous statement to be true, then a model of military mechanization must be in
some sense a cognitive model. Some agent or set of agents perceives the outside world;

\subsection{The utility of agent-based modeling}

Good for insurgency because its hidden, and

In the typology of Axtell \citep{axtell2000agents}

ABM is useful for exactly such settings.

@@
